Spatial Reasoning and Data Displays

### 2015-03-28 Paper/InfoVis2015/TestingVisualAptitude.Rnw
intro

% \usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}

@@

%---------------------------------------------------
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{color}
% \usepackage[dvipsnames,svgnames]{xcolor}
% \usepackage{ulem}
% \usepackage[section]{placeins}
\usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
% \usepackage{sidecap}
\usepackage{multirow}
\usepackage{bbm}
\usepackage{url}
\usepackage{xr}
\usepackage{xr-hyper}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Susan VanderPlas, and Heike Hofmann, \textit{Member, IEEE}}
\authorfooter{
%% insert punctuation at end of each item
\item Susan VanderPlas is a graduate student at Iowa State University. Email: skoons@iastate.edu.
\item Heike Hofmann is a professor at Iowa State University in the Departments of Statistics and Human Computer Interaction. Email: hofmann@iastate.edu.
}

@@

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
    \includegraphics[width=.4\linewidth]{lineup}
    \caption[Sample Lineup]{Sample lineup of boxplots. Participants are instructed to choose the plot which appears most different from the others. In this lineup, plot 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}}
  }

@@

``Lineups" have recently been introduced \cite{buja2009statistical, wickham2010graphical, majumder2013validation} as a tool to evaluate the statistical significance of a graphical finding. Lineups are also useful in assessing the effectiveness of different graphical displays \cite{hofmann2012graphical, loy:2015}. Like their police counterpart, lineups consist of several distractor plots (of randomly generated data) and one target (the data plot). 
% 
% \begin{figure}[htp]
% 
% \end{figure}

@@

Assume that an observer has evaluated $K$ lineups of size $m$ (consisting of $m-1$ decoy plots and 1 target), and successfully identified the target in $n_c$ of these plots, while missing the target in $n_w$ of them. 
The score for this individual is then given as:
\begin{eqnarray}\label{eq.scoring1}
n_c - n_w/(m-1).
\end{eqnarray}
Note that the sum of answers, $n_c + n_w$, is at most $K$, but may be less, if an observer chooses to not answer one of the lineup tasks or runs out of time.

@@

\begin{figure}[ht]
\includegraphics[width=.9\linewidth]{Figure/fig-ResultsScaledScores-1}
\caption{Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \label{fig:Scores}}
\end{figure}
%

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}

@@

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}
loadfonts()
require(grid)
ggplot()+
  geom_point(data=subset(all, points), aes(x=x, y=y), shape=1) + 
  geom_smooth(data=subset(all, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=subset(all, !points), aes(x=x, y=y, label=paste0("Correlation = \n", label))) + 
  geom_line(data=densities, aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank(),
        plot.margin=unit(c(0,0,0,0), unit="mm"))
rm(list=c("data", "all", "densities"))
embedFonts("./Figure/fig-VisReasoningSPM-1.pdf")
@
\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}

>>>>

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}

@@

%---------------------------------------------------
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
\usepackage{color}
% \usepackage[dvipsnames,svgnames]{xcolor}
% \usepackage{ulem}
% \usepackage[section]{placeins}
\usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
% \usepackage{sidecap}
\usepackage{multirow}
\usepackage{bbm}
\usepackage{url}
\usepackage{xr}
\usepackage{xr-hyper}
\usepackage{afterpage}

\usepackage{todonotes}
\newcommand{\hhnote}[1]{\todo[inline,color=magenta!40]{#1}}

@@

%% indicate IEEE Member or Student Member in form indicated below
\author{Susan VanderPlas, and Heike Hofmann, \textit{Member, IEEE}}
\authorfooter{
%% insert punctuation at end of each item
\item Susan VanderPlas is a graduate student at Iowa State University. Email: skoons@iastate.edu.
\item Heike Hofmann is Professor of Statistics at Iowa State University and member of the faculty in the Human Computer Interaction program. Email: hofmann@iastate.edu.
}

@@

%% Uncomment below to include a teaser figure.
   \teaser{
    \centering
    \includegraphics[width=.4\linewidth]{lineup}
    \caption[Sample Lineup]{``Which of these plots is the most different from the others?"\newline
    This is the question participants are asked to answer for each of the lineups in the study. In the figure, a sample lineup of boxplots is shown. %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\label{fig:lineup}}
  }

@@

In this paper we investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol  \cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

``Lineups" have recently been introduced as a tool to evaluate the statistical significance of a graphical finding. Lineups are also useful in assessing the effectiveness of different graphical displays \cite{hofmann2012graphical, loy:2015}. Like their police counterpart, lineups consist of several distractor plots (of randomly generated data) and one target (the data plot). 
% 
% \begin{figure}[htp]
% 
% \end{figure}

@@

Assume that an observer has evaluated $K$ lineups of size $m$ (consisting of $m-1$ decoy plots and 1 target), and successfully identified the target in $n_c$ of these plots, while missing the target in $n_w$ of them. 
The score for this individual is then given as:
\begin{eqnarray}\label{eq.scoring1}
n_c - \frac{n_w}{m-1}.
\end{eqnarray}
Note that the sum of answers, $n_c + n_w$, is at most $K$, but may be less, if an observer chooses to not answer one of the lineup tasks or runs out of time.

@@

\begin{figure}[ht]
\includegraphics[width=.9\linewidth]{Figure/fig-ResultsScaledScores-1}
\caption{Test scores for lineups and visuospatial tests. As none of the participants scored at or below zero, we can conclude that there is little evidence of random guessing. We also note the score compression that occurs on the Visual Search test; this indicates that most participants scored extremely high, and thus, participants' scores are not entirely representative of their ability. \label{fig:Scores}}
\end{figure}
% 
\afterpage{\clearpage}

@@

qplot(data=lineup.summary.categorical, x=value, y=lineup, geom="boxplot", fill=I("grey")) + 
  facet_wrap(~variable, scales="free_x", ncol=4) + xlab("") + 
  ylab("Scaled Lineup Score") + 
  geom_point(aes(x=value, y=lineup), shape=1, size=3) + 
  theme_bw() + 
  theme(plot.margin=unit(c(0,0,0,0), unit="mm"))
embedFonts("./Figure/fig-VisReasoningCategorical-1.pdf")
@
\begin{figure*}[h!tb]\centering
\includegraphics[width=.85\linewidth]{fig-VisReasoningCategorical-1}
\caption[Visual Aptitude Study Results]{Demographic characteristics of participants compared with lineup score. Categories are ordered by effect size; majoring in a STEM field, calculus completion, hours spent playing video games per week, and sex are all associated with a significant difference in lineup score. }\label{fig:visualaptitudecat}
\end{figure*}
\afterpage{\clearpage}

@@

  rbind.fill(
    data.frame(
      xvar = names(data)[ycol], 
      yvar = names(data)[xcol],
      x = data[, xcol], 
      y = data[, ycol], 
      points = T,
      data
    ),
    data.frame(
      yvar = names(data)[ycol], 
      xvar = names(data)[xcol],
      x = mean(range(data[, ycol])), 
      y = mean(range(data[, xcol])),
      label = round(cor(data[,xcol], data[,ycol]), 3),
      points = F
    )
  )
}))
all$xvar <- factor(all$xvar, levels=names(data))
all$yvar <- factor(all$yvar, levels=names(data))
densities <- do.call("rbind", lapply(1:ncol(data), function(i) {
    tmp <- density(data[,i], cut = 1, adjust=.75)
    data.frame(
      xvar = names(data)[i], 
      yvar = names(data)[i],
      x = tmp$x, 
      y = tmp$y/max(tmp$y)* diff(range(tmp$x)) + min(tmp$x)
    )
  }))
fix.names <- function(values){
  values %>% 
    str_replace("lineup", "Lineups") %>% 
    str_replace("vis_search", "Visual Search") %>%
    str_replace("fig_class", "Figure Classification") %>% 
    str_replace("card_rot", "Card Rotation") %>%
    str_replace("folding", "Paper Folding")
}
loadfonts()
require(grid)
ggplot()+
  geom_point(data=subset(all, points), aes(x=x, y=y), shape=1) + 
  geom_smooth(data=subset(all, points), aes(x=x, y=y), method="lm") + 
  geom_text(data=subset(all, !points), aes(x=x, y=y, label=paste0("Correlation = \n", label))) + 
  geom_line(data=densities, aes(x=x, y=y)) + 
  facet_grid(xvar~yvar, scales="free", labeller = labeller(xvar=fix.names, yvar=fix.names)) + 
  theme_bw() + 
  theme(axis.title=element_blank(),
        plot.margin=unit(c(0,0,0,0), unit="mm"))
rm(list=c("data", "all", "densities"))
embedFonts("./Figure/fig-VisReasoningSPM-1.pdf")
@
\begin{figure}[ht]
\includegraphics[width=\linewidth]{Figure/fig-VisReasoningSPM-1}
\caption{Pairwise scatterplots of test scores. Lineup scores are most highly correlated with figure classification scores, and are also highly correlated with card rotation scores. Paper folding and card rotation scores are also highly correlated.\label{fig:scatterplotmatrix}}
\end{figure}
\afterpage{\clearpage}

@@

<<modelresults, echo=FALSE, results='asis'>>=
library(xtable)
print(xtable(summary(m1), caption="Estimates and significances of a linear model of lineup scores.", label="tab:m1"), caption.placement="top")
@
\afterpage{\clearpage}
%
\subsection{Lineup Types}
Each of the three sets of 20 lineups was taken from previous experiments examining which type of plot most effectively conveyed important characteristics of the underlying data set. 
\begin{figure*}[htbp]\centering
\subfloat[Set 1: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/boxplot_1_15_5_4_13}}
\subfloat[Set 1: Density plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/density_1_15_5_4_13}}
\subfloat[Set 1: Dotplot ]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/dotplot_1_15_5_4_13}}
\subfloat[Set 1: Histogram]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups1/histogram_1_15_5_4_13}}\\
\subfloat[Set 2: Boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f94f6965d7}}
\subfloat[Set 2: Boxplot with jittered points]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9e65f0e4}}
\subfloat[Set 2: Bee swarm boxplot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f9fc58d73}}
\subfloat[Set 2: Violin plot]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups2/file147f93acbb7e4}}\\
\subfloat[Set 3: QQ-plot with guide line]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13876a225d34-single}}
\subfloat[Set 3: QQ-plot with acceptance bands]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13877c93dc78-single}}
\subfloat[Set 3: QQ-plot rotated 45 degrees.]{
  \includegraphics[keepaspectratio=T,width=.245\linewidth]{../../Images/Lineups3/file13878e5c92f-single}}
\caption{Lineup plot types used in the experiment. \label{fig:lineups}}
\end{figure*}
